Predator clusters

### 2023-08-10 reports/04_pred_clus.qmd
refactoring cluster report

tar_load(c("predators_stations", "predators_abundant", 
           "seaice_conc_df", "seaice_contours",
           "sightings_clust", "sightings_mtx", "sightings_dist", 
           "zoop_sf", "ant_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
seaice_contours <- terra::unwrap(seaice_contours)
```

@@

Dietrich et al. (2021) restricted cluster analysis to taxa observed at 5% or more of stations, which was necessary to avoid giving rare species too much leverage. Predators are far less diverse, so this analysis retains all species except uncertain IDs (e.g. UNSE = unknown seal). `r n_distinct(predators_abundant$species)` species included in analysis.

@@

total_stations <- n_distinct(zoop_sf$amlr.station)
pred_frac <- predators_abundant %>%
  as_tibble() %>% 
  group_by(amlr.station, species) %>% 
  summarize(count_station = sum(count),
            .groups = "drop") %>% 
  group_by(species) %>%
  summarize(total_count = sum(count_station),
            stations_present = n(),
            station_frac = stations_present / total_stations) %>%
  arrange(desc(station_frac))

@@

We use the gap statistic [@tibshirani2001] to choose the optimal number of clusters ($k$). The gap statistic ($f$) is a goodness of clustering measure. The authors recommended choosing $k$ at the shoulder. Heuristically, the shoulder is the smallest $k$ such that $f(k) \geq f(k+1) - s(k+1)$ where $s$ is the standard error of $f$. According to this rule, the predators are best described by one cluster. However, there's a clear jump from $k=2$ to $k=3$, followed by a leveling off. Therefore, based on our prior knowledge of three predator guilds (pagophilic, open water, and marginal ice), we'll present the results for $k=3$ here.

@@

cluster_fn <- function(x, k) {
  bray_dist <- function(x) {
    vegdist(x, method = "bray")
  }
  list(cluster = cutree(hclust(bray_dist(x), method = "ward.D2"), k = k))
}
set.seed(139)
sightings_gap <- suppressWarnings(
  clusGap(sightings_mtx, 
                   FUN = cluster_fn,
                   K.max = 10, 
                   B = 500,
                   verbose = FALSE)
)
sightings_gap_df <- as_tibble(sightings_gap$Tab) %>% 
  mutate(
    k = seq_along(logW),
    is_optimum = k == maxSE(gap, SE.sim, method = "Tibs2001SEmax")
  )
ktibs <- filter(sightings_gap_df, is_optimum) %>% pull(k)
tibs_thr <- with(sightings_gap_df, gap[ktibs + 1] - SE.sim[ktibs + 1])
ggplot(sightings_gap_df, aes(x = k, y = gap)) +
  geom_line() +
  geom_pointrange(aes(ymin = gap - SE.sim, ymax = gap + SE.sim,
                      color = is_optimum)) +
  geom_segment(x = ktibs, xend = ktibs + 1, 
               y = tibs_thr, yend = tibs_thr,
               color = "blue",
               linetype = "dashed") +
  scale_color_manual(values = c(`TRUE` = "blue",
                                `FALSE` = "black")) +
  theme_classic() +
  theme(legend.position = "none")
```

@@

# Cut predators into three clusters
clust <- cutree(sightings_clust, 3)
g <- split(names(clust), clust)
p <- ggtree(sightings_clust, hang = -1)
pred_clust_mrca <- sapply(g, function(n) MRCA(p, n))

@@

How do the predator clusters compare to the zooplankton clusters?

```{r}
#| label: predator_mosaic
#| warning: false

# Join with zooplankton clusters
both_clust <- zoop_sf %>% 
  transmute(amlr.station, 
            zoop_clust = factor(`Winter Cluster factor`)) %>% 
  right_join(tibble(amlr.station = names(clust),
                    pred_clust = factor(clust)),
             by = "amlr.station")

# Mosaic plot
cluster_pal = c(`1` = "#96B687",
                `2a` = "#9CB2C8",
                `2b` = "#CAB3CC",
                `3a` = "#D29A97",
                `3b` = "#E8AD5C")
ggplot(both_clust) +
  geom_mosaic(aes(x = product(zoop_clust, pred_clust), 
                  fill = zoop_clust)) +
  geom_mosaic_text(aes(x = product(zoop_clust, pred_clust),
                       label = after_stat(.wt)),
                   as.label = TRUE) +
  scale_fill_manual(values = cluster_pal) +
  labs(x = "Predator cluster",
       y = "Zooplankton cluster",
       fill = "Zooplankton cluster") +
  theme_classic()
```

The above mosaic plot shows the distribution of stations by predator and zooplankton clusters. For example, the bottom left rectangle indicates 34 stations were assigned to zooplankton cluster 1 and predator cluster 1. Rectangles color-coded by zooplankton cluster using the color palette from Dietrich et al. 2021 Fig 3a.

Upon visual inspection of the mosaic plot, there appears to be some general correlation. Predator cluster 1 associates somewhat indiscriminately with zooplankton clusters 2a-3b. Predator cluster 2 associates strongly with zooplankton cluster 2a, as well as with 2b to a lesser extent. Predator cluster 3 associates with zooplankton clusters 3a and 3b. Zooplankton cluster 1 seems to be the "background".

For context, zooplankton cluster 1 was mostly salps, clusters 2a and 2b were small krill and amphipods (mostly offshore), and clusters 3a and 3b were large krill and *Metridia* (mostly in the Bransfield Strait).

@@

TODO: add text about how to interpret values.

@@

species_wide <- predators_abundant %>% 
  select(amlr.station, species, count_nmi) %>% 
  pivot_wider(names_from = species,
              values_from = count_nmi,
              values_fill = 0) %>% 
  column_to_rownames("amlr.station")
cluster_indval <- labdsv::indval(species_wide, clust)
indval_mtx <- cluster_indval$indval
indval_fmt <- apply(indval_mtx, 2, function(x) {
  ord <- order(x, decreasing = TRUE)
  result <- ifelse(x >= 0.25, 
         str_glue("{rownames(indval_mtx)} ({round(x * 100)}%)"),
         NA) %>% 
    `[`(ord) %>% 
    na.omit() %>% 
    paste(collapse = "<br>")
  ifelse(result == "", "None", result)
})
as_tibble(t(indval_fmt)) %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling()

@@

# Adelie penguin x cluster 3 worked example
predators_abundant2 <- predators_abundant %>% 
  mutate(cluster = map_int(amlr.station, \(x) clust[x]))
adpn <- filter(predators_abundant2, species == "ADPN")

pic_adpn <- sum(adpn$cluster == 3)
nc3 <- sum(clust == 3)
fic_adpn <- pic_adpn / nc3

abu_adpn <- adpn %>% 
  group_by(cluster) %>% 
  summarize(abu = sum(count_nmi) / n())
aic_adpn <- abu_adpn$abu[3] / sum(abu_adpn$abu)

# ANFU x cluster 1
anfu <- filter(predators_abundant2, species == "ANFU")

pic_anfu <- sum(anfu$cluster == 1)
nc1 <- sum(clust == 1)
fic_anfu <- pic_anfu / nc1

abu_anfu <- anfu %>% 
  group_by(cluster) %>% 
  summarize(abu = sum(count_nmi) / n())
aic_anfu <- abu_anfu$abu[1] / sum(abu_anfu$abu)

@@

### Species within clusters

```{r}
#| label: clusters_species
predators_abundant %>% left_join(
  tibble(
    amlr.station = names(clust),
    pred_clust = factor(clust,
                        labels = c("Open", "Marginal", "Pagophilic"))
  ), by = "amlr.station") %>% 
  complete(amlr.station, pred_clust, species, 
           fill = list(count_nmi = 0)) %>% 
  group_by(species, pred_clust) %>% 
  summarize(mean_abundance = mean(count_nmi), .groups = "drop") %>%
  mutate(species = fct_reorder(species, 
                               mean_abundance, 
                               .fun = sum,
                               .desc = TRUE)) %>% 
  ggplot(aes(species, mean_abundance)) +
  geom_col() +
  labs(y = "Mean abundance (ind nmi^-1)") +
  facet_wrap(~ pred_clust, ncol = 1) +
  theme_classic(base_size = 14) +
  theme(axis.title.x = element_blank())
```

@@

station_ice <- predators_abundant %>%
  distinct(amlr.station, ice_coverage)

pred_clust <- tibble(
  amlr.station = names(clust),
  pred_clust = factor(clust,
                      labels = c("Open", "Marginal", "Pagophilic"))
) %>% 
  left_join(select(zoop_sf,
                   amlr.station,
                   geometry,
                   Year),
            by = "amlr.station") %>% 
  left_join(station_ice, by = "amlr.station")

pred_clust %>% 
  mutate(Year = factor(Year)) %>% 
  crosstable::crosstable("Year", by = "pred_clust", total = "col")

@@

pred_clust_sf <- pred_clust %>% 
  st_as_sf() %>% 
  st_transform(ant_proj())

map_lim <- st_bbox(pred_clust_sf) %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.2)

icy_open <- filter(pred_clust_sf,
                   pred_clust == "Open",
                   ice_coverage >= 7.5)

ant_basemap(map_lim) +
  geom_sf(aes(color = pred_clust), 
          pred_clust_sf, 
          size = 2, 
          shape = 15) +
  geom_sf(data = seaice_contours, color = "skyblue") +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_gradient(lim = c(0, 1), na.value = "transparent") +
  coord_ant(map_lim) +
  theme(legend.position = "bottom")
```

@@

nmds_stress <- tibble(
  k = 1:6,
  stress = map_dbl(k, function(k) {
    metaMDS(
      sightings_mtx, 
      distance = "bray",
      k = k,
      trymax = 100,
      trace = FALSE
    )$stress
  })
)

@@

sightings_nmds <- metaMDS(
  sightings_mtx, 
  distance = "bray",
  k = 3,
  trymax = 100,
  trace = FALSE
)

nmds_isoreg <- isoreg(sightings_nmds$diss, sightings_nmds$dist)

nmds_shepard <- tibble(
  diss = sightings_nmds$diss,
  dist = sightings_nmds$dist,
  mono = nmds_isoreg$yf
)

nmds_spear <- cor(nmds_shepard$diss,
                  nmds_shepard$dist,
                  method = "spearman")
nmds_label <- str_glue(
  "Non-metric fit, R^2 = {round(1 - sightings_nmds$stress^2, 3)}\n",
  "Linear fit, R^2 = {round(nmds_spear^2, 3)}\n",
  "k = {sightings_nmds$ndim}\n",
  "Stress = {round(sightings_nmds$stress, 3)}"
)

@@

nmds_df <- sightings_nmds$points %>% 
  as_tibble(rownames = "amlr.station") %>% 
  rename_with(~ paste0("N", .x), starts_with("MDS")) %>% 
  right_join(pred_clust_sf %>% 
               as_tibble() %>% 
               select(amlr.station, pred_clust),
             by = "amlr.station")

@@

@@

# Evironmental variables
station_env <- zoop_sf %>% 
  select(amlr.station, zuml_m, avg.temp, avg.salinity, Integ.chla.100m, 
         Integ.phae.100m, TOD_2levels_civil, Year, 
         zoop = `Winter Cluster factor`) %>% 
  as_tibble() %>% 
  left_join(station_ice, by = "amlr.station") %>% 
  mutate(Year = factor(Year)) %>% 
  slice(map_int(rownames(sightings_mtx),
                ~ which(.x == amlr.station))) %>% 
  select(-amlr.station, -geometry)

# Fit multiple regression model
pred_envfit <- envfit(sightings_nmds, 
                      station_env, 
                      permutations = 999, 
                      choices = 1:3,
                      na.rm = TRUE)

@@

```{r}
#| label: env_summ

env <- tibble(
  amlr.station = names(clust),
  pred_clust = factor(clust,
                      labels = c("Open", "Marginal", "Pagophilic"))
) %>% 
  left_join(
    select(as_tibble(zoop_sf),
           amlr.station, avg.salinity, Integ.chla.100m),
    by = "amlr.station") %>% 
  left_join(select(station_ice, amlr.station, ice_coverage),
            by = "amlr.station") %>% 
  pivot_longer(avg.salinity:ice_coverage, 
               names_to = "envvar",
               values_to = "envval")
env %>% 
  group_by(envvar) %>% 
  summarize(envaov = list(aov(envval ~ pred_clust))) %>% 
  mutate(Fval = map_dbl(envaov, \(aov) summary(aov)[[1]]$`F value`[1]),
         pval = map_dbl(envaov, \(aov) summary(aov)[[1]]$`Pr(>F)`[1]))
env %>% 
  group_by(pred_clust, envvar) %>% 
  summarize(across(envval, list(mean = partial(mean, na.rm = TRUE),
                                sd = partial(sd, na.rm = TRUE))),
            .groups = "drop")
```

@@

```{r}
pred_clust_ice <- tibble(
  amlr.station = names(clust),
  pred_clust = factor(clust,
                      labels = c("Open", "Marginal", "Pagophilic"))
) %>% 
  left_join(station_ice,
            by = "amlr.station") %>% 
  drop_na(ice_coverage)

@@

```{r}
pred_prey <- as_tibble(pred_clust) %>% 
  select(amlr.station, pred_clust) %>% 
  left_join(select(as_tibble(zoop_sf), 
                   amlr.station, 
                   prey_clust = `Winter Cluster factor`),
            by = "amlr.station")
pred_prey_chisq <- with(pred_prey, chisq.test(pred_clust, prey_clust))
pred_prey_chisq
pred_prey_chisq$residuals

>>>>

tar_load(
  c("predators_clust", "stations_clust", "sightings_cut",
    "seaice_conc_df", "seaice_contours", "stations_ice",
    starts_with("sightings"),
    starts_with("nmds"),
    "zoop_sf", "ant_sf"), 
  store = here::here("_targets")
)
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
seaice_contours <- terra::unwrap(seaice_contours)
```

@@

Dietrich et al. (2021) restricted cluster analysis to taxa observed at 5% or more of stations, which was necessary to avoid giving rare species too much leverage. Predators are far less diverse, so this analysis retains all species except uncertain IDs (e.g. UNSE = unknown seal). `r n_distinct(predators_clust$species)` species included in analysis.

@@

total_stations <- n_distinct(zoop_sf$amlr.station)
pred_frac <- predators_clust %>%
  as_tibble() %>% 
  group_by(amlr.station, species) %>% 
  summarize(count_station = sum(count),
            .groups = "drop") %>% 
  group_by(species) %>%
  summarize(total_count = sum(count_station),
            stations_present = n(),
            station_frac = stations_present / total_stations) %>%
  arrange(desc(station_frac))

@@

We use the gap statistic [@tibshirani2001] to choose the optimal number of clusters ($k$). The gap statistic ($f$) is a goodness of clustering measure. The authors recommended choosing $k$ at the shoulder of the $f \sim k$ curve. Heuristically, the shoulder is the smallest $k$ such that $f(k) \geq f(k+1) - s(k+1)$ where $s$ is the standard error of $f$. According to this rule, the predators are best described by one cluster. However, there's a clear jump from $k=2$ to $k=3$ (red), followed by a leveling off. Therefore, based on our prior knowledge of three predator guilds (pagophilic, open water, and marginal ice), we'll present the results for $k=3$ here.

@@

ktibs <- with(sightings_gap, k[is_optimum])
tibs_thr <- with(sightings_gap, gap[ktibs + 1] - SE.sim[ktibs + 1])
ggplot(sightings_gap, aes(x = k, y = gap)) +
  geom_line() +
  geom_pointrange(aes(ymin = gap - SE.sim, ymax = gap + SE.sim,
                      color = is_optimum)) +
  geom_pointrange(aes(ymin = gap - SE.sim, ymax = gap + SE.sim),
                  filter(sightings_gap, k == 3),
                  color = "red") +
  geom_segment(x = ktibs, xend = ktibs + 1, 
               y = tibs_thr, yend = tibs_thr,
               color = "blue",
               linetype = "dashed") +
  scale_color_manual(values = c(`TRUE` = "blue",
                                `FALSE` = "black")) +
  theme_classic() +
  theme(legend.position = "none")
```

@@

g <- split(names(sightings_cut), sightings_cut)
p <- ggtree(sightings_clust, hang = -1)
pred_clust_mrca <- sapply(g, function(n) MRCA(p, n))

@@

@@

@@

indval_mtx <- sightings_indval$indval
indval_fmt <- apply(indval_mtx, 2, function(x) {
  ord <- order(x, decreasing = TRUE)
  result <- ifelse(x >= 0.25, 
         str_glue("{rownames(indval_mtx)} ({round(x * 100)}%)"),
         NA) %>% 
    `[`(ord) %>% 
    na.omit() %>% 
    paste(collapse = "<br>")
  ifelse(result == "", "None", result)
})
as_tibble(t(indval_fmt)) %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling()

@@

# Adelie penguin x cluster 3 worked example
adpn <- filter(predators_clust, species == "ADPN")

pic_adpn <- sum(adpn$pred_clust == "Pagophilic")
nc3 <- sum(sightings_cut == 3)
fic_adpn <- pic_adpn / nc3

abu_adpn <- adpn %>% 
  group_by(pred_clust) %>% 
  summarize(abu = sum(count_nmi) / n())
aic_adpn <- abu_adpn$abu[3] / sum(abu_adpn$abu)

# ANFU x cluster 1
anfu <- filter(predators_clust, species == "ANFU")

pic_anfu <- sum(anfu$pred_clust == "Open")
nc1 <- sum(sightings_cut == 1)
fic_anfu <- pic_anfu / nc1

abu_anfu <- anfu %>% 
  group_by(pred_clust) %>% 
  summarize(abu = sum(log(count_nmi + 1)) / n())
aic_anfu <- abu_anfu$abu[1] / sum(abu_anfu$abu)

@@

@@

stations_clust %>% 
  mutate(Year = factor(Year)) %>% 
  crosstable::crosstable("Year", by = "pred_clust", total = "col") %>% 
  crosstable::as_flextable()

@@

map_lim <- st_bbox(stations_clust) %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.2)

icy_open <- filter(stations_clust,
                   pred_clust == "Open",
                   ice_coverage >= 7.5)

ant_basemap(map_lim) +
  geom_sf(aes(color = pred_clust), 
          stations_clust, 
          size = 2, 
          shape = 15) +
  # geom_sf(data = filter(stations_clust, 
  #                       amlr.station %in% c("AMLR2015_W1111",
  #                                           "AMLR2015_W1113")),
  #         color = "yellow") +
  geom_sf(data = seaice_contours, color = "skyblue") +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_gradient(lim = c(0, 1), na.value = "transparent") +
  coord_ant(map_lim) +
  theme(legend.position = "bottom")
```

@@

@@

nmds_spear <- cor(nmds_shepard$diss,
                  nmds_shepard$dist,
                  method = "spearman")
nmds_label <- str_glue(
  "Non-metric fit, R^2 = {round(1 - nmds_sightings$stress^2, 3)}\n",
  "Linear fit, R^2 = {round(nmds_spear^2, 3)}\n",
  "k = {nmds_sightings$ndim}\n",
  "Stress = {round(nmds_sightings$stress, 3)}"
)

@@

@@

```{r}
#| label: env-loadings-old
#| eval: false

@@

# Evironmental variables
station_env <- zoop_sf %>% 
  select(amlr.station, zuml_m, avg.temp, avg.salinity, Integ.chla.100m, 
         Integ.phae.100m, TOD_2levels_civil, Year, 
         zoop = `Winter Cluster factor`) %>% 
  as_tibble() %>% 
  left_join(stations_ice, by = "amlr.station") %>% 
  mutate(Year = factor(Year)) %>% 
  slice(map_int(rownames(sightings_mtx),
                ~ which(.x == amlr.station))) %>% 
  select(-amlr.station, -geometry)

# Fit multiple regression model
pred_envfit <- envfit(nmds_sightings, 
                      station_env, 
                      permutations = 999, 
                      choices = 1:3,
                      na.rm = TRUE)

@@

```{r}
#| label: env_summ
#| eval: false

env <- tibble(
  amlr.station = names(clust),
  pred_clust = factor(clust,
                      labels = c("Open", "Marginal", "Pagophilic"))
) %>% 
  left_join(
    select(as_tibble(zoop_sf),
           amlr.station, avg.salinity, Integ.chla.100m),
    by = "amlr.station") %>% 
  left_join(select(stations_ice, amlr.station, ice_coverage),
            by = "amlr.station") %>% 
  pivot_longer(avg.salinity:ice_coverage, 
               names_to = "envvar",
               values_to = "envval")
env %>% 
  group_by(envvar) %>% 
  summarize(envaov = list(aov(envval ~ pred_clust))) %>% 
  mutate(Fval = map_dbl(envaov, \(aov) summary(aov)[[1]]$`F value`[1]),
         pval = map_dbl(envaov, \(aov) summary(aov)[[1]]$`Pr(>F)`[1]))
env %>% 
  group_by(pred_clust, envvar) %>% 
  summarize(across(envval, list(mean = partial(mean, na.rm = TRUE),
                                sd = partial(sd, na.rm = TRUE))),
            .groups = "drop")
```

@@

```{r}
#| eval: false
pred_clust_ice <- tibble(
  amlr.station = names(clust),
  pred_clust = factor(clust,
                      labels = c("Open", "Marginal", "Pagophilic"))
) %>% 
  left_join(stations_ice,
            by = "amlr.station") %>% 
  drop_na(ice_coverage)

@@

```{r}
#| eval: false
pred_prey <- as_tibble(pred_clust) %>% 
  select(amlr.station, pred_clust) %>% 
  left_join(select(as_tibble(zoop_sf), 
                   amlr.station, 
                   prey_clust = `Winter Cluster factor`),
            by = "amlr.station")
pred_prey_chisq <- with(pred_prey, chisq.test(pred_clust, prey_clust))
pred_prey_chisq
pred_prey_chisq$residuals
